On Extending a Full-Sharing Multithreaded Tabling Design with Batched Scheduling

### 2015-05-26 paper.tex
paper -> revising 1

\begin{abstract}
Tabling is an implementation technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Multithreading in Prolog is the
ability to concurrently perform computations, in which each
computation runs independently but shares the program clauses. When
multithreading is combined with tabling, we have the best of both
worlds, since we can exploit the combination of higher procedural
control with higher declarative semantics. To support this
combination, the Yap Prolog system has, at engine level, multiple
designs that vary from a No-Sharing design, where each thread
allocates fully private tables for each new tabled subgoal called
during its computation, to a Full-Sharing design, where threads share
the complete table space. Arguably, batched scheduling is one of the
most successful tabling scheduling strategies. In this work, we
propose an extension of the Full-Sharing design to support batched
scheduling. Experimental results show that, despite the extra data
structures required to support the extension, the FS design remains
quite competitive.\\

@@

Batched scheduling favors forward execution first, backtracking next,
and consuming answers or completion last. It thus tries to delay the
need to move around the search tree by batching the return of answers
to repeated subgoals. When new answers are found for a particular
tabled subgoal, they are added to the table space and the evaluation
continues. Batched scheduling can be an useful strategy in tabled
logic programs that require an eager propagation of answers and/or do
not require the complete set of answers to be found (partial
evaluation of the search space).

@@

\begin{wrapfigure}{r}{4.5cm}
%\vspace{-\intextsep}
\centering
\includegraphics[width=4.5cm]{figures/table-space.pdf}
\caption{Table space organization}
\label{fig_table_space}
%\vspace{-0.5\intextsep}
\end{wrapfigure}

@@

In Yap, a thread views its tables as private but, at the engine level,
it implements three designs for concurrent tabling support that vary
from a \emph{No-Sharing} (NS) design, where each thread allocates
fully private tables for each new tabled subgoal called during its
computation, to a \emph{Full-Sharing} (FS) design, where threads share
the complete table space. Figure~\ref{fig_table_space} shows Yap's
multithreaded table space organization for the NS and FS designs,
where an interface layer abstracts the design being used at the engine
level. The figure illustrates the main differences between the two
designs for a situation where several threads are evaluating the same
tabled subgoal call $call\_i$.

@@

We now present experimental results about the usage of PAC in the FS
design with batched scheduling. Since without PAC the FS design would
not be able to be used with batched scheduling, to put results in
perspective, we will be showing also the results for local scheduling
and for the NS design.

The environment for our experiments was a machine with 32-Core AMD
Opteron (TM) Processor 6274 (2 sockets with 16 cores each) with 32G of
main memory, running the Linux kernel is the 3.16.7-200.fc20.x86\_64
with Yap Prolog 6.3. For the experiments, we used the \emph{TabMalloc}
memory allocator~\cite{Areias-12b} and the set of benchmarks described
in~\cite{Areias-12b}. These benchmarks create \emph{worst case
  scenarios}, where we are able to show the lowest bounds of
performance that each design might achieve when applied/used in other
real world applications/programs.

@@

By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
thread, on the minimum overhead ratio, local scheduling is better on
the NS design, while batched scheduling is better in the FS
design. For the average and maximum overhead ratios, in the FS design,
the best strategy is always the local scheduling. Additionally, the
difference between local and batched scheduling in the FS is higher
than in the NS design, which indicates the overhead that the PAC
strategy brings to the FS design. Recall, whenever an answer is found
during the evaluation, PAC requires threads to traverse their private
consumer data structures to check if the answer was already
propagated. This is the reason for overhead that we are observing in
the experimental results.

As we scale the number of threads, for the NS design both scheduling
strategies have similar minimum, average and maximum overhead
ratios. For the FS design, the best minimum overhead ratio is always
for the batched scheduling. For the average and maximum overhead
ratio, local scheduling is always better than batched
scheduling. However, for the average ratio, the results between both
strategies is quite close. For the maximum overhead ratio, the
difference between both strategies is significantly higher than in the
NS design.

In resume, the experimental results show that on average, the PAC
strategy does not seem to have a big impact in the performance,
however it still leaves room for further improvements, because the
difference between local and batched scheduling is higher in the FS
design than in the NS design.

@@

Local and batched scheduling are arguably two of the most well-known
tabling scheduling techniques. The major difference between them is
that local propagates answers only after all answer are found, while
batched propagates answers immediately after they are found. Batched
scheduling is a useful strategy in tabled logic programs that require
an eager propagation of answers and/or do not require the complete set
of answers to be found (partial evaluation of the search space).

In this work, we have presented the PAC strategy, which is a simple
and novel approach for combining the FS design with batched
scheduling. PAC splits answer representation from answer propagation,
and allows the first to be publicly shared among threads while the
second to be private to each thread.

@@

>>>>

\begin{abstract}
Tabling is an implementation technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Multithreading in Prolog is the
ability to concurrently perform computations, in which each
computation runs independently but shares the program clauses. When
multithreading is combined with tabling, we have the best of both
worlds, since we can exploit the combination of higher procedural
control with higher declarative semantics. To support this
combination, the Yap Prolog system has, at engine level, multiple
designs that vary from a No-Sharing design, where each thread
allocates fully private tables for each new tabled subgoal called
during its computation, to a Full-Sharing design, where threads share
the complete table space. Arguably, batched scheduling is one of the
most successful tabling scheduling strategies. In this work, we
propose the PAC strategy to combine the Full-Sharing design with
batched scheduling. Experimental results show that, PAC is a good
first approach, since with it the FS design remains quite
competitive.\\

@@

Batched scheduling favors forward execution first, backtracking next,
and consuming answers or completion last. It thus tries to delay the
need to move around the search tree by batching the return of answers
to repeated subgoals. When new answers are found for a particular
tabled subgoal, they are added to the table space and the evaluation
continues. Batched scheduling can be an useful strategy in tabled
logic programs that require an eager propagation of answers and/or do
not require the complete set of answers to be found.

@@

\begin{wrapfigure}{r}{4.5cm}
\vspace{-\intextsep}
\centering
\includegraphics[width=4.5cm]{figures/table-space.pdf}
\caption{Table space organization}
\label{fig_table_space}
\vspace{-0.5\intextsep}
\end{wrapfigure}

@@

In Yap, a thread views its tables as private but, at the engine level,
it implements three designs for concurrent tabling support that vary
from a \emph{No-Sharing} (NS) design, where each thread allocates
fully private tables for each new tabled subgoal called during its
computation, to a \emph{Full-Sharing} (FS) design, where threads share
the complete table space. Figure~\ref{fig_yap_mt_support} shows Yap's
multithreaded table space organization for the NS and FS designs,
where an interface layer abstracts the design being used at the engine
level. The figure illustrates the main differences between the two
designs for a situation where several threads are evaluating the same
tabled subgoal call $call\_i$.

@@

We now present experimental results about the usage of PAC in the FS
design with batched scheduling. Since without PAC the FS design would
not be able to be used with batched scheduling, to put results in
perspective, we will be showing also the results for local scheduling
and for the NS design. The environment for our experiments was a
machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets with
16 cores each) with 32G of main memory, running the Linux kernel is
the 3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments,
we used the \emph{TabMalloc} memory allocator~\cite{Areias-12b} and
the set of benchmarks described in~\cite{Areias-12b}. These benchmarks
create \emph{worst case scenarios}, where we are able to show the
lowest bounds of performance that each design might achieve when
applied/used in other real world applications/programs.

@@

By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
thread, on the minimum overhead ratio, local scheduling is better on
the NS design, while batched scheduling is better in the FS
design. For the average and maximum overhead ratios, in the FS design,
the best strategy is always the local scheduling. Additionally, the
difference between local and batched scheduling in the FS is higher
than in the NS design, which indicates the overhead that the PAC
strategy brings to the FS design. Recall, whenever an answer is found
during the evaluation, PAC requires threads to traverse their private
consumer data structures to check if the answer was already
propagated. This is the reason for difference in the overheads that we
are observing in the experimental results. As we scale the number of
threads, for the NS design both scheduling strategies have similar
minimum, average and maximum overhead ratios. For the FS design, the
best minimum overhead ratio is always for the batched scheduling. For
the average and maximum overhead ratio, local scheduling is always
better than batched scheduling. However, for the average ratio, the
results between both strategies is quite close. For the maximum
overhead ratio, the difference between both strategies is
significantly higher than in the NS design. In conclusion,
experimental results show that on average, the PAC strategy does not
seem to have a big impact in the performance, however it still leaves
room for further improvements. The difference between both local and
batched scheduling is significantly higher in the FS design than in
the NS design.

@@

Local and batched scheduling are arguably two of the most well-known
tabling scheduling techniques. The major difference between them is
that local propagates answers only after all answer are found, while
batched propagates answers immediately after they are found. Batched
scheduling is a useful strategy in tabled logic programs that require
an eager propagation of answers and/or do not require the complete set
of answers to be found (partial evaluation of the search space). In
this work, we have presented the PAC strategy, which is a simple and
novel approach for combining the FS design with batched
scheduling. PAC splits answer representation from answer propagation,
and allows the first to be publicly shared among threads while the
second to be private to each thread.

@@

This work is funded by the project SIBILA (NORTE-07-0124-FEDER-000059)
under the North Portugal Regional Operational Programme (ON.2 – O Novo
Norte) and under the National Strategic Reference Framework (NSRF),
through the European Regional Development Fund (ERDF), and by national
funds, through the Portuguese funding agency, Fundação para a Ciência
e a Tecnologia (FCT).
